---
title: "data_table_tutorial"
output: html_document
date: "2024-06-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# data.table

## data.table intro


dt[row,col,group]

### import library
```{r}
library("data.table")
library("magrittr")
```

### import data
```{r}
df <- read.csv("https://raw.githubusercontent.com/jinseob2kim/lecture-snuhlab/master/data/example_g1e.csv")

# doesn't have to be a csv file
dt <- fread("https://raw.githubusercontent.com/jinseob2kim/lecture-snuhlab/master/data/example_g1e.csv")

class(df);class(dt)
```


import only necessary data due to memory and speed issues

```{r}
names(dt)


#dt <- fread("https://raw.githubusercontent.com/jinseob2kim/lecture-snuhlab/master/data/example_g1e.csv", select=)

dt1 <- fread("https://raw.githubusercontent.com/jinseob2kim/lecture-snuhlab/master/data/example_g1e.csv", select=c("EXMD_BZ_YYYY", "RN_INDI", "HME_YYYYMM"))
head(dt1)

dt2 <- fread("https://raw.githubusercontent.com/jinseob2kim/lecture-snuhlab/master/data/example_g1e.csv", select = 1:5)
head(dt2)

dt3 <- fread("https://raw.githubusercontent.com/jinseob2kim/lecture-snuhlab/master/data/example_g1e.csv", drop=6:10)
head(dt3)


```

### write data (save)

```{r}

write.csv(dt, "data.table.csv", row.names=F) 

#recommended
fwrite(dt, "datatable_fwrite.csv")
```

cf. fst is the fastest method

### row operation
```{r}
dt[1:10] # dt[1:10,] is the same
head(dt, 10)

dt[(EXMD_BZ_YYYY %in% 2009:2012) & (BMI >=25)]

#data.frame way
df[(df$EXMD_BZ_YYYY %in% 2009:2012) & (df$BMI >=25),]
subset(df, (EXMD_BZ_YYYY %in% 2009:2012) & (BMI>=25))
with(df, df[(EXMD_BZ_YYYY %in% 2009:2012) & (BMI>=25),])
# with is default in data.table


dt[order(HME_YYYYMM)]

dt[order(HME_YYYYMM,HGHT)]
dt[order(HME_YYYYMM, -HGHT)]


```

once start with pipe, can't use [] again
```{r}

dt[(EXMD_BZ_YYYY %in% 2009:2012) & (BMI >= 25)][order(HGHT)][(order(GFR))]

#same
dt[(EXMD_BZ_YYYY %in% 2009:2012) & (BMI >= 25)] %>% 
  .[order(HGHT)] %>% 
  .[order(GFR)]

temp.dt <- dt[(EXMD_BZ_YYYY %in% 2009:2012) & (BMI >= 25)][order(HGHT)][(order(GFR))]

temp.dt1 <- dt[(EXMD_BZ_YYYY %in% 2009:2012) & (BMI >= 25)]
temp.dt1

temp.dt2 <- temp.dt1[order(HGHT)]
temp.dt2

temp.dt3 <- temp.dt2[order(GFR)]

all.equal(temp.dt, temp.dt3)
all.equal(temp.dt2, temp.dt3)


```

## column

```{r}

dt[, 1:10]
dt[, c("HGHT","WGHT")]

dt[, .(HGHT, WGHT)]
# same
dt[, list(HGHT, WGHT)] # .() == list()

#same but rename col
dt[, .(Height = HGHT, Weight=WGHT)]

#return vector
dt[, HGHT]


colvars <- grep("Q_", names(dt), value=T)


#cf with data.frame
df[, colvars]

dt[, ..colvars] # confusing
dt[, colvars, with=F]
dt[, .SD, .SDcols = colvars]

dt[, .SD, .SDcols = -colvars]
```
 
## column summary

```{r}
dt[, .(mean(HGHT), mean(WGHT), mean(BMI))]
dt[, .(HGHT = mean(HGHT), WGHT = mean(WGHT), BMI = mean(BMI))] # rename col

dt[, lapply(.SD, mean), .SDcols = c("HGHT","WGHT","BMI")]

dt[, lapply(.SD, mean)] # NA for missing data
```


```{r}
dt[HGHT >= 175, .N, by=.(EXMD_BZ_YYYY, Q_SMK_YN)]
dt[HGHT >= 175, .N, by =c("EXMD_BZ_YYYY","Q_SMK_YN")]

dt[HGHT >= 175, .N, keyby= c("EXMD_BZ_YYYY", "Q_SMK_YN")] # keep the order
dt[HGHT >= 175, .N, keyby= .(EXMD_BZ_YYYY, Q_SMK_YN)]

```

keyby includes NA values. Therefore, we need to remove the NAs in advance

```{r}
dt[HGHT >= 175]

dim(dt[HGHT >= 175])

newdt <- dt[HGHT >=175] %>% na.omit()
dim(newdt)
# newdt2 <- dt[HGHT >=175] %>% na.omit.data.table()
# dim(newdt2)

# this is wrong because output is null
newdt1 <- dt[HGHT >=175, .N] %>% na.omit()
dim(newdt1)

temp.dt <- dt[HGHT >=175] %>% na.omit() %>% 
  .[, .N, keyby=c("EXMD_BZ_YYYY", "Q_SMK_YN")]

temp.dt
dim(temp.dt)


dt[HGHT >=175] %>% na.omit() %>% 
  .[, .N, keyby= .(EXMD_BZ_YYYY >= 2015, Q_SMK_YN == 1)]

# same
dt[HGHT >=175] %>% na.omit() %>% 
  .[, .N, keyby=.(get("EXMD_BZ_YYYY") >= 2015, get("Q_SMK_YN") == 1)]

dt[HGHT >= 175, .N, keyby= .(Y2015 = ifelse(EXMD_BZ_YYYY >= 2015, ">=2015", "<2015"))]
```

get() : find string in the data.table. If it is not in the data.table, find the one as global variable


## new variables

```{r}

dt[, BMI2 := round(WGHT/(HGHT/100)^2, 1)]
#same
dt$BMI2 <- round(dt$WGHT/(dt$HGHT/100)^2, 1)


# multiple variables using `:=`
dt[, `:=`(BP_SYS140 = factor(as.integer(BP_SYS >= 140)), BMI25 = factor(as.integer(BMI >= 25)))]

# remove col
dt[, BMI2 := NULL]

# .SD : all cols

# first medical record
dt[order(EXMD_BZ_YYYY), .SD[1], keyby="RN_INDI"]
# same
dt[order(EXMD_BZ_YYYY), lapply(.SD, head,1), keyby="RN_INDI"]


# last medical record
dt[order(EXMD_BZ_YYYY), .SD[.N], keyby="RN_INDI"]



# dt[order(EXMD_BZ_YYYY), .SD, keyby="RN_INDI"]
# dim(dt[order(EXMD_BZ_YYYY), .SD, keyby="RN_INDI"])
# dim(dt[order(EXMD_BZ_YYYY), .SD[1], keyby="RN_INDI"])
```


### specific symbol
```{r}

```



### merge

```{r}
colvars

dt1 <- dt[1:10, .SD, .SDcols = c("EXMD_BZ_YYYY", "RN_INDI", "HME_YYYYMM", colvars)] # only q
dt2 <- dt[6:15, -..colvars] # everything but q

# 6-10 are in both data.table


merge(dt1, dt2, by = c("EXMD_BZ_YYYY", "RN_INDI", "HME_YYYYMM"), all = T)    # Full join (union)
merge(dt1, dt2, by = c("EXMD_BZ_YYYY", "RN_INDI", "HME_YYYYMM"), all = F)    # Inner join (intersection)
merge(dt1, dt2, by = c("EXMD_BZ_YYYY", "RN_INDI", "HME_YYYYMM"), all.x = T)  # left join
merge(dt1, dt2, by = c("EXMD_BZ_YYYY", "RN_INDI", "HME_YYYYMM"), all.y = T)  # right join

# difference (anti join)

# dt[criteria, ]

dt2[dt1, on = c("EXMD_BZ_YYYY", "RN_INDI", "HME_YYYYMM")]  # left join2
dt1[dt2, on = c("EXMD_BZ_YYYY", "RN_INDI", "HME_YYYYMM")]  # right join2

dt1[!dt2, on = c("EXMD_BZ_YYYY", "RN_INDI", "HME_YYYYMM")] # left anti join
dt2[!dt1, on = c("EXMD_BZ_YYYY", "RN_INDI", "HME_YYYYMM")] # right anti join


```


## melt

```{r}
names(dt)
```


wide to long
```{r}

# basic with warnings
dt.long1 <- melt(dt, id.vars = c("EXMD_BZ_YYYY", "RN_INDI", "HME_YYYYMM"))

dt.long1 <- melt(dt, 
                 id.vars = c("EXMD_BZ_YYYY", "RN_INDI", "HME_YYYYMM"), # fixed
                 measure.vars = c("TOT_CHOL", "TG", "HDL", "LDL") # vars to modify
                 
                 )

dt.long1

dt.long1 <- melt(dt, 
                 id.vars = c("EXMD_BZ_YYYY", "RN_INDI", "HME_YYYYMM"), # fixed
                 measure.vars = c("TOT_CHOL", "TG", "HDL", "LDL"), # vars to modify
                 variable.name= "Lipid", # new variable name
                 value.name = "value" # name of the new value for each variable
                 )

dt.long1

```

## enhanced melt

multiple groups
```{r}
col1 <- c("BP_SYS", "BP_DIA") # Systolic: 수축 Diastolic: 이완
col2 <- c("VA_LT", "VA_RT") # left and right

dt.long2 <- melt(dt,
                 id.vars = c("EXMD_BZ_YYYY", "RN_INDI", "HME_YYYYMM"),
                 measure = list(col1, col2),
                 value.name = c("BP","VA"))

dt.long2

dt.long2[, variable := factor(variable, labels = c("SBP/VA_LT", "DBP/VA_RT"))] %>% select(variable="DBP/VA_RT")

dt.long2

```



## decast
long to wide
```{r}


```


## data.table file


```{r}
library(haven);library(data.table);library(magrittr);library(fst)
# Set core number when data.table
setDTthreads(0)  ## 0: All

## SAS to fst/csv
for (v in c("bnc", "bnd", "m20", "m30", "m40", "m60", "inst", "g1e_0208", "g1e_0915")){
  read_sas(file.path("data", paste0("nsc2_", v, "_1000.sas7bdat"))) %>% 
    #write_fst(file.path("data", paste0("nsc2_", v, "_1000.fst")))
    fwrite(file.path("data", paste0("nsc2_", v, "_1000.csv")))
}


## fst
inst <- read_fst("data/nsc2_inst_1000.fst", as.data.table = T)
bnc <- read_fst("data/nsc2_bnc_1000.fst", as.data.table = T) 
bnd <- read_fst("data/nsc2_bnd_1000.fst", as.data.table = T) 
m20 <- read_fst("data/nsc2_m20_1000.fst", as.data.table = T) 
m30 <- read_fst("data/nsc2_m30_1000.fst", as.data.table = T) 
m40 <- read_fst("data/nsc2_m40_1000.fst", as.data.table = T) 
m60 <- read_fst("data/nsc2_m60_1000.fst", as.data.table = T) 
g1e_0915 <- read_fst("data/nsc2_g1e_0915_1000.fst", as.data.table = T) 


## csv
inst <- fread("data/nsc2_inst_1000.csv")
bnc <- fread("data/nsc2_bnc_1000.csv") 
bnd <- fread("data/nsc2_bnd_1000.csv") 
m20 <- fread("data/nsc2_m20_1000.csv") 
m30 <- fread("data/nsc2_m30_1000.csv") 
m40 <- fread("data/nsc2_m40_1000.csv") 
m60 <- fread("data/nsc2_m60_1000.csv") 
g1e_0915 <- fread("data/nsc2_g1e_0915_1000.csv") 
```


## 

```{r}
library(data.table);library(magrittr)
# Set core number when data.table
setDTthreads(0)  ## 0: All

## csv
inst <- fread("data/nsc2_inst_1000.csv")
bnc <- fread("data/nsc2_bnc_1000.csv") 

## Death date: last day in month
bnd <- fread("data/nsc2_bnd_1000.csv")[, Deathdate := (lubridate::ym(DTH_YYYYMM) %>% lubridate::ceiling_date(unit = "month") - 1)][]
m20 <- fread("data/nsc2_m20_1000.csv") 
m30 <- fread("data/nsc2_m30_1000.csv") 
m40 <- fread("data/nsc2_m40_1000.csv")[SICK_CLSF_TYPE %in% c(1, 2, NA)]            ## Exclude 3
m60 <- fread("data/nsc2_m60_1000.csv") 
g1e_0915 <- fread("data/nsc2_g1e_0915_1000.csv") 



## after 2006, New I10-15 (Hypertensive disease) in Main Sick
# new cases only

# | is or in 
code.HTN <- paste(paste0("I", 10:15), collapse = "|")
data.start <- m20[like(SICK_SYM1, code.HTN) & (MDCARE_STRT_DT >= 20060101), .(Indexdate = min(MDCARE_STRT_DT)), keyby = "RN_INDI"]

## Previous disease: Among all sick code
excl <- m40[(MCEX_SICK_SYM %like% code.HTN) & (MDCARE_STRT_DT < 20060101), .SD[1], .SDcols = c("MDCARE_STRT_DT"), keyby = "RN_INDI"]

## Merge: left anti join
data.incl <- data.start[!excl, on = "RN_INDI"][, Indexdate := as.Date(as.character(Indexdate), format = "%Y%m%d")][]
#data.incl <- data.start[!(RN_INDI %in% excl$RN_INDI)]


## Add age, sex, death
data.asd <- merge(bnd, bnc[, .(SEX = SEX[1]), keyby = "RN_INDI"], by = "RN_INDI") %>% 
  merge(data.incl, by = "RN_INDI") %>% 
  .[, `:=`(Age = year(Indexdate) - as.integer(substr(BTH_YYYY, 1, 4)),
           Death = as.integer(!is.na(DTH_YYYYMM)),
           Day_FU = as.integer(pmin(as.Date("2015-12-31"), Deathdate, na.rm =T) - Indexdate))] %>% .[, -c("BTH_YYYY", "DTH_YYYYMM", "Deathdate")] 




```



